---
title: "Examples with emmeans"
author: "Russ Lenth"
date: "8/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```
The **emmeans** package provides a variety of *post hoc* analyses such
as obtaining estimated marginal means (EMMs) and comparisons thereof,
displaying these results in a graph, and a number of related tasks.

This vignette illustrates basic uses of **emmeans** with `lm_robust` 
objects. For more details, refer to the **emmeans** package itself 
and its vignettes.

### A factorial experiment
The `warpbreaks` dataset provided in base R has the results of a two-factor
experiment. We start by fitting a model
```{r}
library(estimatr)

warp.rlm <- lm_robust(log(breaks) ~ wool * tension, data = warpbreaks)
```
Typical use of `emmeans()` is to obtain predictions, or marginal means thereof,
via a formula of the form `~ primary.variables | by.variables`:
```{r}
library(emmeans)

emm <- emmeans(warp.rlm, ~ tension | wool)
class(emm)
str(emm)
emm
```
These results may be plotted as side-by-side intervals or as an interaction-style plot:
```{r}
plot(emm)
emmip(emm, wool ~ tension, CIs = TRUE)
```

This particular example has a response transformation. That transformation is
detected and we may back-transform to the original scale:
```{r}
confint(emm, type = "response")
```
We may do comparisons and other contrasts:
```{r}
pairs(emm) # pairwise comparisons
contrast(emm, "trt.vs.ctrl", ref = "L", type = "response", adjust = "mvt")
```
Note that with a log transformations, it is possible to back-transform 
comparisons, and they become ratios. With other transformations, back-transforming
is not possible.

### Rank-deficient models
Let's create a variation on this example where one cell is omitted:
```{r}
warpi.rlm <- update(warp.rlm, subset = -(37:48))
(rgi <- ref_grid(warpi.rlm))
summary(rgi)
```
Note that the empty cell is detected and flagged as non-estimable.

Some additional explanation here. EMMs are based on a *reference grid*, defined
as the grid created by all possible combinations of factor levels, together with
the mean of each numerical predictor. The reference grid here (`rgi`) is also an
`"emmGrid"` object just like the previous `emm`.
The grid itself is available as a data frame via the `grid` member, and
you can verify that the above results match those of the `predict` function
for the model:
```{r}
predict(warpi.rlm, newdata = rgi@grid, se.fit = TRUE)
```
There is one exception for the empty cell. I will leave it as a user exercise to demonstrate that if we were to use different contrasts when fitting `warpi.rlm`,
the predictions will be the same *except* for the empty cell.

### Multivariate models
If there is a multivariate response, it is treated as another factor that
is crossed with the other factors in the model. To illustrate, consider
the dataset `MOats`, provided in **emmeans**:
```{r}
MOats.rlm <- lm_robust(yield ~ Block + Variety, data = MOats)
ref_grid(MOats.rlm)
```
By default, the pseudo-factor is named `rep.meas`, but we can change it
if we like:
```{r}
emmeans(MOats.rlm, pairwise ~ nitro, mult.name = "nitro")
```
This illustrates an additional feature of `emmeans` that we can put a contrast method in the left side of a formula.

### Afterword
There are numerous capabilities of **emmeans** not illustrated here. See that package's
help files and vignettes. Using `vignette("basics", "emmeans")` is a good starting point.
