---
title: "estimatr in the tidyverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimatr in the tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 6, fig.height = 3.5)
have <- function(...) all(vapply(c(...), requireNamespace, logical(1), quietly = TRUE))
library(estimatr)
options(digits = 3)
```

Everything estimatr returns is a list, and `tidy()` turns the part you want into a data frame. From there the tidyverse applies unchanged. What follows uses `lm_robust()`, but `tidy()` works the same way on `lm_lin()`, `iv_robust()`, `lh_robust()`, `difference_in_means()` and `horvitz_thompson()` fits.

The examples use the Swiss fertility data that ships with R.

## Getting tidy

```{r}
fit <- lm_robust(Fertility ~ Agriculture + Catholic, data = swiss)
tidy(fit)
```

One row per coefficient, and an `outcome` column, which matters because `lm_robust()` accepts a matrix of outcomes and fits them all at once.

`glance()` gives the one-row model summary instead.

```{r}
glance(fit)
```

## dplyr

```{r, eval = have("dplyr")}
library(dplyr)

fit |>
  tidy() |>
  filter(term != "(Intercept)") |>
  mutate(significant = p.value <= 0.05) |>
  select(term, estimate, std.error, significant)
```

## ggplot2

A coefficient plot is the tidy data frame with `geom_pointrange()` on it.

```{r, eval = have("dplyr", "ggplot2")}
library(ggplot2)

gg_df <- fit |>
  tidy() |>
  filter(term != "(Intercept)")

ggplot(gg_df, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate, with 95% confidence interval", y = NULL) +
  theme_minimal()
```

`geom_smooth()` takes `lm_robust` as a method, so the ribbon reflects the robust variance rather than the classical one.

```{r, eval = have("ggplot2")}
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm_robust", formula = y ~ x) +
  theme_minimal()
```

The formula can be anything `lm_robust()` accepts, polynomials included.

```{r, eval = have("ggplot2")}
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm_robust", formula = y ~ poly(x, 3, raw = TRUE)) +
  theme_minimal()
```

## Many models with purrr

Fitting the same model on subsets, then stacking the results:

```{r, eval = have("dplyr", "purrr")}
library(purrr)

swiss |>
  mutate(educated = Education > 8) |>
  split(~ educated) |>
  map(\(d) lm_robust(Fertility ~ Catholic, data = d)) |>
  map(tidy) |>
  list_rbind(names_to = "educated") |>
  filter(term == "Catholic")
```

Fitting different outcomes on the same regressor:

```{r, eval = have("dplyr", "purrr")}
c("Fertility", "Education", "Agriculture") |>
  map(\(y) lm_robust(reformulate("Catholic", y), data = swiss)) |>
  map(tidy) |>
  list_rbind() |>
  filter(term == "Catholic")
```

Several outcomes at once is also a single call, and considerably faster, since the design matrix is factorized once:

```{r}
tidy(lm_robust(cbind(Fertility, Education, Agriculture) ~ Catholic, data = swiss))
```

## Bootstrapping

`tidy()` output stacks, so a bootstrap is a loop over resamples.

```{r, eval = have("dplyr", "purrr")}
set.seed(1)

boot <- map(1:200, \(i) {
  d <- swiss[sample(nrow(swiss), replace = TRUE), ]
  tidy(lm_robust(Fertility ~ Catholic + Agriculture, data = d))
}) |>
  list_rbind()

boot |>
  group_by(term) |>
  summarise(bootstrap_se = sd(estimate)) |>
  left_join(tidy(lm_robust(Fertility ~ Catholic + Agriculture, data = swiss)) |>
              select(term, hc2_se = std.error), by = "term")
```

The bootstrap standard errors and the HC2 standard errors are close, which is the expected result: both are estimating the same thing without assuming homoskedasticity.
