---
title: "estimatr 2.0: what changes, what does not, and why"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimatr 2.0: what changes, what does not, and why}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
has_estimatr <- requireNamespace("estimatr", quietly = TRUE)
has_randomizr <- requireNamespace("randomizr", quietly = TRUE)
```

## The short version

estimatrZero is a ground-up rewrite of estimatr aimed at the DeclareDesign workflow: fit the same model thousands of times, as fast as possible, and get the same number every time. It is a drop-in replacement for almost everything you already write. Change `estimatr::` to `estimatrZero::` and your code returns identical estimates, identical standard errors, identical confidence intervals, and identical objects, only faster.

Three things are different, and each one is a deliberate trade. `horvitz_thompson()` takes one probability argument instead of five. `fixed_effects` no longer accepts HC2, HC3, or CR2. A handful of stargazer helpers are gone. The rest of this document explains what stays the same, what you gain, and what the three changes buy you.

## Getting it

Do not attach both packages in the same session. They define the same S3 methods for the same classes, and whichever loads second silently takes over dispatch for both. Either attach one:

```r
library(estimatrZero)
lm_robust(y ~ z, data = dat)
```

or, when you genuinely need to compare them, attach neither and qualify every call:

```r
estimatr::lm_robust(y ~ z, data = dat)
estimatrZero::lm_robust(y ~ z, data = dat)
```

## What does not change

The five estimators keep their estimatr signatures exactly:

| function | signature |
|---|---|
| `lm_robust()` | `formula, data, weights, subset, clusters, fixed_effects, se_type, ci, alpha, return_vcov, try_cholesky` |
| `lm_lin()` | `formula, covariates, data, weights, subset, clusters, se_type, ci, alpha, return_vcov, try_cholesky` |
| `iv_robust()` | `formula, data, weights, subset, clusters, fixed_effects, se_type, ci, alpha, diagnostics, return_vcov, try_cholesky` |
| `difference_in_means()` | `formula, data, blocks, clusters, weights, subset, se_type, condition1, condition2, ci, alpha` |
| `lh_robust()` | `..., data, linear_hypothesis` |

Every S3 method estimatr provides is present and behaves the same way: `tidy()`, `glance()`, `summary()`, `print()`, `predict()`, `coef()`, `confint()`, `vcov()`, `nobs()`, `update()`, plus the texreg and emmeans hooks.

```{r}
set.seed(42)
dat <- data.frame(
  y = rnorm(1000), x = rnorm(1000), z = rbinom(1000, 1, 0.5),
  w = runif(1000, 0.5, 1.5), cl = rep(1:100, each = 10),
  bl = rep(1:50, each = 20)
)
```

```{r, eval = has_estimatr}
fit_estimatr <- estimatr::lm_robust(y ~ z + x, data = dat, clusters = cl, weights = w)
fit_zero <- estimatrZero::lm_robust(y ~ z + x, data = dat, clusters = cl, weights = w)

all.equal(fit_estimatr$coefficients, fit_zero$coefficients, tolerance = 1e-12)
all.equal(fit_estimatr$std.error, fit_zero$std.error, tolerance = 1e-12)
all.equal(fit_estimatr$df, fit_zero$df, tolerance = 1e-12)
```

The identity claim is backed by a test suite of 715 passing assertions, 336 of which compare estimatrZero output to estimatr output coefficient by coefficient and standard error by standard error, across every supported SE type, weighted and unweighted, clustered and unclustered, single and multivariate outcomes.

There are exactly two places where the numbers differ on purpose, and both involve `fixed_effects`. They are described below.

## What changes, and why

### 1. `horvitz_thompson()` takes one probability argument

estimatr asks you to say what you know about the randomization in five different ways, and the right combination depends on the design:

```r
# estimatr
horvitz_thompson(Y ~ Z, data = dat, ra_declaration = decl)
horvitz_thompson(Y ~ Z, data = dat, condition_prs = 0.4, simple = TRUE)
horvitz_thompson(Y ~ Z, data = dat, blocks = bl, condition_prs = pr)
horvitz_thompson(Y ~ Z, data = dat, clusters = cl, condition_prs = pr)
horvitz_thompson(Y ~ Z, data = dat, condition_pr_mat = pr_mat)
```

estimatrZero has one argument, `condition_prs`, which takes whichever object you actually have:

```r
# estimatrZero
horvitz_thompson(Y ~ Z, data = dat, condition_prs = decl)                 # ra_declaration
horvitz_thompson(Y ~ Z, data = dat, condition_prs = c("0" = 0.6, "1" = 0.4))  # named vector
horvitz_thompson(Y ~ Z, data = dat, condition_prs = pr_matrix)            # per-unit matrix
```

The `blocks`, `clusters`, `simple`, `condition_pr_mat`, `subset`, and `return_condition_pr_mat` arguments are gone, and `se_type = "constant"` is gone.

**The rationale.** An `ra_declaration` already carries the block structure, the cluster structure, the per-unit marginal probabilities, and whether the randomization was simple or complete. Passing `blocks` and `clusters` separately restates information the declaration already holds, and gives the estimator two sources of truth that can disagree. Any parametric design you can describe in words is a `declare_ra()` call:

```r
declare_ra(blocks = bl, clusters = cl, prob = pi, simple = FALSE)
```

and any design you cannot describe parametrically is a permutation matrix:

```r
declare_ra(permutation_matrix = perm)
```

which replaces `condition_pr_mat` and, in the process, replaces 538 lines of matrix-construction helpers with one `tcrossprod()` call.

**The upside, part one: multi-arm designs.** estimatr refuses an `ra_declaration` with more than two arms. estimatrZero contrasts any two arms of one, with `condition1` and `condition2` picking the contrast. The estimand stays the average treatment effect over all N units of the design, so `data` must carry one row per unit, including the arms outside the contrast; a declaration whose size does not match `nrow(data)` is an error rather than a silent misalignment.

```{r, eval = has_randomizr}
library(randomizr)
set.seed(3)
decl3 <- declare_ra(N = 300, conditions = c("control", "T1", "T2"))
Z3 <- conduct_ra(decl3)
dat3 <- data.frame(Y = rnorm(300) + 0.4 * (Z3 == "T1") + 0.8 * (Z3 == "T2"), Z = Z3)

estimatrZero::horvitz_thompson(Y ~ Z, data = dat3, condition_prs = decl3,
                               condition1 = "control", condition2 = "T2")
```

**The upside, part two: speed.** Passing the declaration is what makes the O(1) variance possible. When estimatrZero knows the design is complete randomization within blocks, it computes the Young's inequality bound from six scalar sums per block rather than from an N-by-N matrix of joint probabilities. Passing a bare probability vector instead still works, and gives you the conservative simple-randomization bound, valid for any design but exact only for Bernoulli assignment. The two paths are now visibly different at the call site: an `ra_declaration` buys you the design-aware variance, and a plain vector buys you the conservative one. In estimatr that distinction was buried in which combination of arguments you happened to supply.

```{r, eval = has_randomizr}
set.seed(2)
decl <- declare_ra(blocks = rep(c("a", "b", "c", "d"), each = 50), prob = 0.4)
Z <- conduct_ra(decl)
dat_ht <- data.frame(Y = rnorm(200) + 0.5 * Z, Z = Z)

estimatrZero::horvitz_thompson(Y ~ Z, data = dat_ht, condition_prs = decl)
```

### 2. `fixed_effects` no longer accepts HC2, HC3, or CR2

In estimatr, `lm_robust(y ~ z, fixed_effects = ~ block)` defaults to HC2 without clusters and CR2 with them. In estimatrZero those requests are an error, and the defaults are `"stata"` (equivalently HC1) without clusters and `"CR0"` with them, which is what `feols` does.

```{r, error = TRUE}
estimatrZero::lm_robust(y ~ z + x, data = dat, fixed_effects = ~ bl,
                        clusters = cl, se_type = "CR2")
```

**The rationale.** HC2, HC3, and CR2 all need the leverage values of the full design matrix, meaning the matrix with every fixed-effect dummy in it. Absorbing fixed effects by demeaning is exactly the decision not to build that matrix. estimatr resolves the conflict by building it anyway, which is why the 2000-block model above takes 43 seconds; the Frisch-Waugh-Lovell demeaning saves nothing if the variance step reconstructs the dummies. The alternative, computing HC2 from the demeaned leverages, would return a different number under the same name. Neither silent slowness nor silent approximation is acceptable, so estimatrZero errors and tells you how to get the number you asked for.

**The upside.** Two things. First, you can absorb a very large number of fixed effects, which was previously impractical. Second, the error message names the escape hatch, so nothing is actually unavailable:

```r
# exactly equal to estimatr's fixed_effects + HC2 result
lm_robust(y ~ z + x + factor(block), data = dat, se_type = "HC2")
```

That equivalence is tested. If you need CR2 with a moderate number of blocks, write the dummies and pay the cost knowingly. If you have thousands of fixed effects, use `fixed_effects` and take CR0.

**One thing to check when you port code.** Because the default changes, a script that fits `lm_robust(y ~ z, fixed_effects = ~ block)` and does not name an `se_type` gets HC2 from estimatr and HC1 from estimatrZero. The point estimates are identical. The standard errors are close, typically within a fraction of a percent without clusters, and CR0 runs a couple of percent below CR2 with clusters, in the usual direction. If your results depend on which one you use, say which one you used.

### 3. Functions that are gone

`starprep()`, `commarobust()`, `declaration_to_condition_pr_mat()`, `gen_pr_matrix_cluster()`, and `permutations_to_condition_pr_mat()` are not exported. The first two are stargazer conveniences; the last three are the condition-probability-matrix machinery that the new Horvitz-Thompson variance does not need. HAC (Newey-West) and CR3 standard errors are also not implemented.

## Porting an existing script

Three greps cover every breaking change. Whatever they do not find needs no attention.

| Grep for | Replace with | Where it bites |
|---|---|---|
| `ra_declaration`, `condition_pr_mat`, `simple =`, or `blocks`/`clusters` inside `horvitz_thompson()` | one `condition_prs =`, holding a declaration, a named vector, or a matrix | Horvitz-Thompson designs |
| `fixed_effects` together with `se_type = "HC2"`, `"HC3"`, or `"CR2"` | `se_type = "stata"` or `"CR0"`, or write the dummies with `factor()` | absorbed fixed effects |
| `starprep`, `commarobust`, `declaration_to_condition_pr_mat`, `gen_pr_matrix_cluster`, `permutations_to_condition_pr_mat` | the first two have no replacement here; the last three are unnecessary | table output, HT internals |

The first two stop with an error naming the replacement. The third stops because the function does not exist.

One change does not announce itself, and it is the one to check by hand: `lm_robust(y ~ z, fixed_effects = ~ block)` with no `se_type` named gets HC2 from estimatr and HC1 from estimatrZero. The point estimates are identical and the standard errors are close, but if a result turns on which one you used, say which one you used.

## What is new

Five things estimatr cannot do, all of which fall out of the rewrite rather than being bolted on.

**Multi-arm Horvitz-Thompson.** estimatr refuses an `ra_declaration` with more than two arms. estimatrZero contrasts any two arms of one, with the estimand still defined over all N units of the design. See [What changes, and why] for the call.

**A joint hypothesis test.** `lh_robust()` returns a `joint_hypothesis` element carrying a Wald F statistic on cluster-adjusted degrees of freedom, which estimatr declines to compute.

**Residuals.** `residuals(fit)` returns them, on the scale of the data, in the original row order for clustered fits, and structural rather than first-stage for `iv_robust()`. In estimatr the slot was `NULL` for every estimator.

**A warning when a regressor is dropped.** Collinear terms come back as NA coefficients in both packages; only one of them tells you which terms it dropped.

**Blocked designs whose blocks are not all the same shape.** estimatr either errors on such a design or applies the matched-pairs estimator to every block, big ones included. estimatrZero uses the Pashley and Miratrix (2021) estimators. See [Blocked designs whose blocks are not all the same shape] for the details, which are worth a section of their own.

## Blocked designs whose blocks are not all the same shape

estimatr handles two kinds of blocked design and refuses everything between them. If every block has at least two treated and two control units, each block carries its own Neyman variance. If every block is a matched pair, the variance comes from the variation across pairs. A design with both, or a block holding one treated unit and three control units, either errors or silently applies the matched-pairs estimator to every block after a warning.

Such designs are not exotic. Coarsened exact matching, full matching, and multisite trials with one or two sites per stratum all produce them.

The substantive correction is to classify blocks by how many units each **arm** holds, not by how large the block is. A block of eight units with one of them treated has no more estimable within-block variance than a matched pair does; the singleton arm gives you one number, and one number has no variance. estimatrZero implements the three estimators of Pashley and Miratrix (2021) on that classification:

- Blocks with at least two units in each arm contribute their own Neyman variance, their equation 4.
- Blocks with a singleton arm contribute through the variation across such blocks, their equation 8, the "unified" estimator, which requires no two blocks to share a size. With equal sizes it reduces to the usual matched-pairs estimator, their equation 5, which is used directly because equation 8 is undefined at two equal-sized blocks.
- A design with both kinds combines the parts by squared share of the sample, their section 3.3.

Here is a design with both: twenty blocks of ten units with five treated, and twelve blocks of four units with one treated.

```{r}
set.seed(7)
big <- data.frame(
  bl = rep(paste0("big", 1:20), each = 10),
  z = rep(rep(0:1, each = 5), times = 20)
)
small <- data.frame(
  bl = rep(paste0("sm", 1:12), each = 4),
  z = rep(c(1, 0, 0, 0), times = 12)
)
dat_bl <- rbind(big, small)
dat_bl$y <- rnorm(nrow(dat_bl)) + 0.3 * dat_bl$z

estimatrZero::difference_in_means(y ~ z, data = dat_bl, blocks = bl)
```

estimatr will not fit that design at all:

```{r, error = TRUE, eval = has_estimatr}
estimatr::difference_in_means(y ~ z, data = dat_bl, blocks = bl)
```

Replace the twelve four-unit blocks with twelve matched pairs and estimatr does fit it, by warning and then treating all thirty-two blocks as pairs. The point estimates agree, because the point estimator was never at issue. The standard error is 37% too large and the degrees of freedom are 31 rather than 167, because twenty blocks that could have supplied their own within-block variance are thrown into the across-block calculation instead.

```{r, eval = has_estimatr, warning = TRUE}
pairs <- data.frame(bl = rep(paste0("pr", 1:12), each = 2), z = rep(c(1, 0), times = 12))
dat_pr <- rbind(big, pairs)
dat_pr$y <- rnorm(nrow(dat_pr)) + 0.3 * dat_pr$z

fit_pr_estimatr <- estimatr::difference_in_means(y ~ z, data = dat_pr, blocks = bl)
fit_pr_zero <- estimatrZero::difference_in_means(y ~ z, data = dat_pr, blocks = bl)

data.frame(
  package = c("estimatr", "estimatrZero"),
  design = c(fit_pr_estimatr$design, fit_pr_zero$design),
  estimate = c(fit_pr_estimatr$coefficients, fit_pr_zero$coefficients),
  std.error = c(fit_pr_estimatr$std.error, fit_pr_zero$std.error),
  df = c(fit_pr_estimatr$df, fit_pr_zero$df)
)
```

The `design` element reports which case applied, so you never have to infer it from the block sizes: `"Blocked"` when every block has two units per arm, `"Matched-pair"` when every block is a pair, `"Small blocks"` when every block has a singleton arm without all being pairs, and `"Hybrid blocked"` when the design mixes the two.

```{r}
estimatrZero::difference_in_means(y ~ z, data = dat_bl, blocks = bl)$design
```

Degrees of freedom are not treated in the paper, which stops at the variance. estimatrZero combines the two components by Welch-Satterthwaite, which reduces to `n - 2K` for an all-big design and to `K - 1` for an all-small one, matching what each literature uses on its own.

Two designs are refused, because the variance genuinely cannot be estimated rather than because the software is unwilling. The first is a design with exactly one block holding a singleton arm, which leaves nothing to compare that block against and would contribute a variance of zero:

```{r, error = TRUE}
dat_one <- rbind(big, data.frame(bl = "sm1", z = c(1, 0, 0, 0)))
dat_one$y <- rnorm(nrow(dat_one))
estimatrZero::difference_in_means(y ~ z, data = dat_one, blocks = bl)
```

The second is a set of differently-sized singleton-arm blocks in which one holds half or more of their units, which is the condition equation 8 needs to stay defined and conservative. Both messages name the offending block and suggest merging blocks or using `lm_robust()` with block fixed effects.

Standard errors agree to 1e-10 with `blkvar::block_estimator(method = "hybrid_p")`, the authors' own implementation, across all-big, all-small, matched-pair, and hybrid designs. All-big designs and matched pairs return exactly what they always did.

**Blocks of clusters are a different problem, and are not covered.** Pashley and Miratrix treat treatment assigned within blocks, not blocks of clusters; clusters appear once in their paper, to be set aside. Blocked designs with `clusters` therefore keep the earlier estimators, and a block with a single treated or single control **cluster** is refused outright. See [Current status] for what turned up when that boundary was checked.

## Speed

Median times on an Apple M4 Pro, R 4.6.0, comparing estimatr 1.0.6 to estimatrZero 0.1.0. Coefficients and standard errors agree to 1e-12 in every row.

| task | estimatr | estimatrZero | speedup |
|---|---:|---:|---:|
| `lm_robust`, HC2, n = 1000 | 0.98 ms | 0.48 ms | 2.0x |
| `lm_robust`, classical, n = 1000 | 0.89 ms | 0.41 ms | 2.1x |
| `lm_robust`, CR2, 100 clusters | 1.26 ms | 0.73 ms | 1.7x |
| `lm_robust`, CR2, weighted | 1.06 ms | 0.82 ms | 1.3x |
| `lm_robust`, CR2, n = 5000, 500 clusters | 5.60 ms | 2.30 ms | 2.4x |
| `lm_robust`, HC2, n = 100 | 0.53 ms | 0.36 ms | 1.5x |
| `lm_lin` | 1.17 ms | 0.65 ms | 1.8x |
| `iv_robust` | 1.23 ms | 0.87 ms | 1.4x |
| `difference_in_means` | 0.75 ms | 0.61 ms | 1.2x |

A simulation that calls `lm_robust()` 10,000 times saves about five seconds at n = 1000. That is the boring, everywhere-at-once gain. The interesting gains show up in the two places where the algorithms actually changed.

**Fixed effects.** estimatrZero absorbs fixed effects by alternating projections in C++, the algorithm `fixest::feols` uses, rather than by building the full design matrix of dummies.

| design | estimatr default | estimatr, HC1 | estimatrZero | speedup vs. default |
|---|---:|---:|---:|---:|
| 50 blocks, n = 1000 | 2.7 ms | 2.0 ms | 0.8 ms | 3.4x |
| 500 blocks, n = 10,000 | 594 ms | 14.3 ms | 2.8 ms | 211x |
| 2000 blocks, n = 40,000 | 43,492 ms | 95.0 ms | 10.9 ms | 3978x |

The default column is the honest comparison, because estimatr defaults to HC2 with fixed effects and HC2 requires the hat matrix of the full design including every dummy. That is why 40,000 observations with 2000 blocks takes estimatr 43 seconds. The HC1 column shows the same computation with a standard error that does not need the full hat matrix, where estimatrZero is still 5x to 9x faster because the demeaning itself is faster.

**Horvitz-Thompson.** estimatr builds the full N-by-N joint inclusion probability matrix. estimatrZero derives the same variance from scalar sufficient statistics, so the complete-randomization case is O(1) in N rather than O(N²).

| design | estimatr | estimatrZero | speedup |
|---|---:|---:|---:|
| complete, N = 200 | 1.5 ms | 0.14 ms | 10x |
| complete, N = 1000 | 13.0 ms | 0.18 ms | 73x |
| complete, N = 3000 | 327 ms | 0.39 ms | 837x |

Standard errors are identical to floating-point precision for simple, complete, blocked, clustered, and blocked-and-clustered designs.

### Reproducing these numbers

```r
library(microbenchmark)
set.seed(42)
n <- 1000
dat <- data.frame(
  y = rnorm(n), x = rnorm(n), z = rbinom(n, 1, 0.5),
  w = runif(n, 0.5, 1.5), cl = rep(1:100, each = 10),
  bl = rep(1:50, each = 20)
)

microbenchmark(
  estimatr = estimatr::lm_robust(y ~ z + x, data = dat, clusters = cl),
  estimatrZero = estimatrZero::lm_robust(y ~ z + x, data = dat, clusters = cl),
  times = 100
)
```

## Why rewrite at all

estimatr is about 6,782 lines; estimatrZero is about 3,377. The reduction is not uniform, and where it came from is the interesting part.

The Horvitz-Thompson variance was the largest saving and the one that looked least likely. estimatr builds the full N-by-N matrix of joint inclusion probabilities, with 538 lines of helpers to construct it for each design family. Young's inequality bound for any complete-randomization design reduces to six scalar sufficient statistics per block, so the matrix never has to exist. Designs that cannot be described parametrically are a permutation matrix, and `tcrossprod()` handles those in one line. The whole estimator is 431 lines, and it is faster by a factor that grows with N.

Fixed effects were the second. Absorbing by alternating projections in C++ is the `feols` algorithm, and it makes the number of fixed effects almost free. The cost is that leverage-based standard errors are no longer available for absorbed models, which is a real restriction and is why estimatrZero errors rather than quietly building the dummy matrix that the absorption was meant to avoid.

The rest is ordinary: one code path where estimatr has several, and the S3 surface generated rather than written out per class.

## The estimatr issue list

All 71 open issues on `DeclareDesign/estimatr` were run against both packages, one reproducer at a time. Where they landed:

| status | n |
|---|---:|
| Fixed in estimatrZero | 26 |
| Feature requests and discussions, unchanged | 23 |
| Out of scope by design, or not actionable | 7 |
| Not reproducible without the reporter's data | 6 |
| Superseded by the rewrite | 5 |
| Still open, real work | 4 |

The fixes worth naming. Blocked designs whose blocks are not all the same shape (#336) now use the Pashley and Miratrix (2021) estimators, where estimatr either errors or applies the matched-pairs estimator to every block; this is the largest single addition. `residuals()` returns something (#345). Rank detection matches `lm()`, so a constant regressor comes back as NA rather than as a coefficient of 1e11 (#351, #395). `predict()` works with fixed effects, with factors, and with no `newdata` (#403, #404). `lh_robust()` uses cluster-adjusted degrees of freedom and returns a joint test (#405, #320, #390). Dropped collinear terms are named rather than silently returned as NA (#411). `augment()` exists, which unlocks the broom-aware packages downstream (#377). Multi-arm Horvitz-Thompson works at all (#183).

The four still open are honest holdouts. Two of them, #412 and #337, ask whether `se_type = "stata"` reproduces Stata's clustered standard errors under fixed effects, and settling that needs Stata rather than more R. #289 needs the reporter to say what remains after the clustered case was fixed. #123 is a running catalogue of S3 methods `lm` has: `residuals` and `variable.names` are now present, `anova` and `simulate` are not.

One defect turned up that has no issue behind it, and it is worth knowing if you run block-clustered designs. A block holding a single treated or single control **cluster** has no estimable within-block variance, and both packages returned one anyway. Exhaustive enumeration puts the estimate at 0.12 to 0.25 of the truth, depending on the block's size, for 88% coverage of a nominal 95% interval. estimatrZero now refuses those designs and names the offending blocks. Matched-pair clustered designs are unaffected, since their variance comes from across blocks rather than within them.

## Current status

estimatrZero is not on CRAN yet, and it is not yet used in a published analysis. The test suite is 715 assertions across 230 blocks, `R CMD check` is clean, and the numerical comparisons against estimatr run across 50 seeds per design type for every Horvitz-Thompson design family.

What follows is the list of places where estimatrZero deliberately returns a different number from estimatr. These are behavior changes too, in the sense that estimatrZero returns a different number than estimatr in the cases below. In each one the estimatr number is wrong.

**`lh_robust()` used residual degrees of freedom with clustered models.** With n = 100 and 10 clusters, `lh_robust()` tested hypotheses against 97 degrees of freedom rather than the cluster-adjusted 9, producing confidence intervals far too narrow and inconsistent with the `lm_robust()` fit it was built from. estimatrZero looks up the per-coefficient degrees of freedom, and takes the conservative minimum for composite hypotheses. (estimatr issue #405.)

**`lh_robust()` reported no joint test.** estimatrZero returns a `joint_hypothesis` element with a Wald F statistic on the same cluster-adjusted degrees of freedom. (estimatr issue #320.)

**`iv_robust(diagnostics = TRUE)` returned a first-stage F test with no p-value**, and computed the Wu-Hausman numerator degrees of freedom from the column count of the first-stage residual matrix, which overcounts by one because the intercept residuals are collinear and get dropped. estimatrZero uses the actual rank increase.

**Ordered factors could not be used as clusters.** `class(x) %in% c("factor", "integer")` returns a length-2 logical for an ordered factor, which `if()` rejects. (estimatr issue #421.)

**A formula stored in a variable could not be passed to `fixed_effects`.** The quosure captured the variable name rather than the formula. (estimatr issue #348.)

**`lm_robust(y ~ 1, fixed_effects = ~ block)` crashed** after accumulating 50 convergence warnings on a zero-column design matrix. estimatrZero returns a well-formed intercept-only result with the correct residual degrees of freedom. (estimatr issue #303.)

**Residuals were not returned at all.** `residuals(fit)` was `NULL` for every estimator, though `fitted.values` was present. estimatrZero returns residuals on the scale of the data, in the original row order for clustered fits, and structural rather than first-stage residuals for `iv_robust()`. (estimatr issue #345.)

```{r}
fit <- estimatrZero::lm_robust(y ~ z + x, data = dat)
head(round(residuals(fit), 4))
all.equal(residuals(fit) + fitted(fit), dat$y, check.attributes = FALSE)
```

**A collinear regressor was dropped silently.** The dropped term comes back as an NA coefficient with no warning, which is what made an `lm()` user open the issue after finding coefficients that disagreed between the two functions. estimatrZero warns and names the terms it dropped. (estimatr issue #411.)

```{r, warning = TRUE}
dat$x_copy <- dat$x
fit <- estimatrZero::lm_robust(y ~ x + x_copy, data = dat)
```

**A joint hypothesis test errored.** `lh_robust(Y ~ X1 + X2 + X3, linear_hypothesis = c("X1", "X2"))` should return one joint test; estimatr replies that it "implements tests for hypotheses involving linear combinations of variables but not joint hypotheses." The `joint_hypothesis` element above covers it. (estimatr issue #390.)

**An all-`NA` outcome errored when `weights` were given**, though the same model without weights returned an NA coefficient. The asymmetry bites when one model is fitted across many subgroups and one subgroup has no observed outcome. (estimatr issue #370.)

**Still open.** HAC (Newey-West) and CR3 standard errors are not implemented, and neither is `se_type = "constant"` for Horvitz-Thompson. The multi-arm Horvitz-Thompson contrast is ours to defend rather than a port: estimatr refuses the case outright, so there is no reference implementation to check against. It is conservative in every design simulated so far, with the estimated standard error running 1.08 to 1.32 times the true sampling standard deviation, but nobody has published the bound for a two-of-K contrast.
