---
title: "estimatr in the Tidyverse"
author: "Shikhar Singh"
output:
  html_document:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
library(knitr)
```

`estimatr` is for (fast) OLS and IV regression with robust standard errors. This document shows how `estimatr` integrates with RStudio's `tidyverse` suite of packages.

We use the Swiss Fertility and Socioeconomic Indicators data (available in <tt>R</tt>, description [here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/swiss.html)) to show how `lm_robust` works with `broom`, `dplyr`, `ggplot2`, and `purrr`. What is shown for `lm_robust` here typically applies to all the other `estimatr` functions (`lm_robust`, `difference_in_mean`, `lm_lin`, `iv_robust`, and `horovitz_thompson`).

# Getting tidy

The first step to the tidyverse is turning model output into data we can manipulate. The `tidy` function converts an `lm_robust` object into a data.frame.

```{r, tidy = T, message = F, warning=T}
library(estimatr)
fit <- lm_robust(Fertility ~ Agriculture + Catholic, data = swiss)
tidy(fit)
```

# Data manipulation with `dplyr` 

Once a regression fit is a data.frame, you can use any of the `dplyr` "verbs" for data manipulation, like `mutate`,`filter`, `select`, `summarise`, `group_by`, and `arrange` (more on this [here](https://dplyr.tidyverse.org)). 

```{r, tidy = TRUE, message = FALSE, warning = FALSE}
library(tidyverse)

# lm_robust and filter
fit %>% tidy %>% filter(term == "Agriculture")

# lm_robust and select
fit %>% tidy %>% select(term, estimate, std.error)

# lm_robust and mutate
fit %>% tidy %>% mutate(t_stat = estimate/ std.error,
                        significant = p.value <= 0.05)
```

# Data visualization with `ggplot2` 

`ggplot2` offers a number of data visualization tools that are compatible with `estimatr`

1. Make a coefficient plot:

```{r, tidy = T, message=F, warning=F}
fit %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>%
  ggplot(aes(y = term, x = estimate)) + 
  geom_vline(xintercept = 0, linetype = 2) + 
  geom_point() + 
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, height = 0.1)) + 
  theme_bw()
```

2. Put CIs based on robust variance estimates (rather than the "classical" variance estimates) with the `geom_smooth` and `stat_smooth` functions. 

```{r, tidy = T, message=F, warning=F}
library(ggplot2)
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point() +
  geom_smooth(method = "lm_robust") +
  theme_bw()
```

Note that the functional form can include polynomials. For instance, if the model is $Fertility \sim Agriculture + Agriculture^2 + Agriculture^3$, we can model this in the following way:  

```{r, tidy = T, message=F, warning=F}
library(ggplot2)
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point() +
  geom_smooth(method = "lm_robust",
              formula = y ~ poly(x, 3, raw = TRUE)) +
  theme_bw()
```

# Bootstrap using `rsample`

The `rsample` pacakage provides tools for bootstrapping:

```{r, tidy =T, message = F, warning=F}
library(rsample)

boot_out <-
  bootstraps(data = swiss, 500)$splits %>%
  map(~ lm_robust(Fertility ~ Catholic + Agriculture, data = analysis(.))) %>% 
  map(tidy) %>%
  bind_rows(.id = "bootstrap_replicate")
kable(head(boot_out))
```

`boot_out` is a data.frame that contains estimates from each boostrapped sample. We can then use `dplyr` functions to summarize the bootstraps, `tidyr` functions to reshape the estimates, and `GGally::ggpairs` to visualize them.

```{r, tidy =T, message = F, warning=F}
boot_out %>%
  group_by(term) %>%
  summarise(boot_se = sd(estimate))

# To visualize the sampling distribution

library(GGally)
boot_out %>% 
  select(bootstrap_replicate, term, estimate) %>%
  spread(key = term, value = estimate) %>%
  select(-bootstrap_replicate) %>%
  ggpairs(lower = list(continuous = wrap("points", alpha = 0.1))) +
  theme_bw()
```

# Multiple models using `purrr`

`purrr` provides tools to perform the same operation on every element of a vector. For instance, we may want to estimate a model on different subsets of data. We can use the `map` function to do this.

```{r}
library(purrr)

# Running the same model for highly educated and less educated cantons/districts

two_subsets <- 
  swiss %>%
  mutate(HighlyEducated = as.numeric(Education > 8)) %>%
  split(.$HighlyEducated) %>%
  map( ~ lm_robust(Fertility ~ Catholic, data = .)) %>%
  map(tidy) %>%
  bind_rows(.id = "HighlyEducated")

kable(two_subsets, digits =2)
```

Alternatively, we might want to regress different dependent variables on the same independent variable. `map` can be used alongwith `estimatr` functions for this purpose as well. 

```{r, tidy = T, message=F, warning=F}
three_outcomes <-
  c("Fertility", "Education", "Agriculture") %>%
  map(~ formula(paste0(., " ~ Catholic"))) %>%
  map(~ lm_robust(., data = swiss)) %>%
  map_df(tidy)

kable(three_outcomes, digits =2)
```

Using `ggplot2`, we can make a coefficient plot:

```{r, tidy= T}

three_outcomes %>%
  filter(term == "Catholic") %>%
  ggplot(aes(x = estimate, y = outcome)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point() + 
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, height = 0.1)) + 
  ggtitle("Slopes with respect to `Catholic`") + 
  theme_bw()
```

# Concluding thoughts

Using `estimatr` functions in the tidyverse is easy once the model outputs have been turned into data.frames. We accomplish this with the `tidy` function.  After that, so many summary and visualization possibilities open up. Happy tidying!


