---
title: "estimatr in the tidyverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimatr in the tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 6, fig.height = 3.5)
have <- function(...) all(vapply(c(...), requireNamespace, logical(1), quietly = TRUE))
library(estimatr)
options(digits = 3, width = 90)
```

Everything estimatr returns is a list, and `tidy()` turns the part you want into a tibble. From there the tidyverse applies unchanged. What follows uses `lm_robust()`, but `tidy()` works the same way on `lm_lin()`, `iv_robust()`, `lh_robust()`, `difference_in_means()` and `horvitz_thompson()` fits.

The examples use the Swiss fertility data that ships with R: 47 French-speaking provinces in 1888, with a standardized fertility measure and, as percentages, the share of men in agriculture, the share of Catholics, and the share educated beyond primary school.

```{r}
head(swiss[, c("Fertility", "Agriculture", "Catholic", "Education")], 3)
```

## Getting tidy

```{r}
fit <- lm_robust(Fertility ~ Agriculture + Catholic, data = swiss)
tidy(fit)
```

One row per coefficient, and an `outcome` column, which matters because `lm_robust()` accepts a matrix of outcomes and fits them all at once.

`glance()` gives the one-row model summary instead.

```{r}
glance(fit)
```

`augment()` returns the data the model was fit on, one row per observation, with the fitted values and residuals added as `.fitted` and `.resid`. With `newdata`, it predicts on new rows instead.

```{r}
augment(fit)

augment(fit, newdata = data.frame(Agriculture = c(20, 60), Catholic = c(10, 90)))
```

## dplyr

```{r, eval = have("dplyr")}
library(dplyr)

fit |>
  tidy() |>
  filter(term != "(Intercept)") |>
  mutate(significant = p.value <= 0.05) |>
  select(term, estimate, std.error, significant)
```

## ggplot2

A coefficient plot is the tidy tibble with `geom_pointrange()` on it.

```{r, eval = have("dplyr", "ggplot2")}
library(ggplot2)

gg_df <- fit |>
  tidy() |>
  filter(term != "(Intercept)")

ggplot(gg_df, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate, with 95% confidence interval", y = NULL) +
  theme_minimal()
```

`geom_smooth()` takes `lm_robust` as a method, so the ribbon reflects the robust variance rather than the classical one.

```{r, eval = have("ggplot2")}
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm_robust", formula = y ~ x) +
  theme_minimal()
```

The formula can be anything `lm_robust()` accepts, polynomials included.

```{r, eval = have("ggplot2")}
ggplot(swiss, aes(x = Agriculture, y = Fertility)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm_robust", formula = y ~ poly(x, 3, raw = TRUE)) +
  theme_minimal()
```

A residual plot starts from `augment()`.

```{r, eval = have("ggplot2")}
gg_df <- augment(fit)

ggplot(gg_df, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.6) +
  labs(x = "Fitted values", y = "Residuals") +
  theme_minimal()
```

## Many models with purrr

Fitting the same model on subsets, then stacking the results:

```{r, eval = have("dplyr", "purrr")}
library(purrr)

swiss |>
  mutate(educated = Education > 8) |>
  split(~ educated) |>
  map(\(d) lm_robust(Fertility ~ Catholic, data = d)) |>
  map(tidy) |>
  list_rbind(names_to = "educated") |>
  filter(term == "Catholic")
```

For grouped fits, dplyr's `reframe()` does the same without leaving the data frame. `pick(everything())` hands each group's rows to `lm_robust()`.

```{r, eval = have("dplyr")}
swiss |>
  as_tibble() |>
  mutate(educated = Education > 8) |>
  reframe(tidy(lm_robust(Fertility ~ Catholic, data = pick(everything()))), .by = educated) |>
  filter(term == "Catholic")
```

Fitting different outcomes on the same regressor is the same move after a reshape: stack the outcomes long, nest by outcome, and fit within each.

```{r, eval = have("dplyr", "purrr", "tidyr")}
library(tidyr)

swiss |>
  pivot_longer(c(Fertility, Education, Agriculture),
               names_to = "outcome_variable",
               values_to = "Y") |>
  nest(.by = outcome_variable) |>
  mutate(fit = map(data, \(d) tidy(lm_robust(Y ~ Catholic, data = d)))) |>
  unnest(fit)
```

`lm_robust()` also accepts several outcomes in one call using `cbind`, which is faster because the design matrix is factorized once. Be careful if missingness differs across outcomes because the `cbind` triggers listwise deletion.

```{r, eval = have("dplyr")}
lm_robust(cbind(Fertility, Education, Agriculture) ~ Catholic, data = swiss) |>
  tidy()
```


## Bootstrapping

[rsample](https://rsample.tidymodels.org), from tidymodels, draws the resamples: `bootstraps()` returns a tibble with one resample per row, and `analysis()` extracts each one's data. Because `tidy()` output stacks, the rest is a `map()` and an `unnest()`.

```{r, eval = have("dplyr", "purrr", "tidyr", "rsample")}
library(rsample)

set.seed(343)

boot <- bootstraps(swiss, times = 200) |>
  mutate(coefs = map(splits, \(split) {
    analysis(split) |>
      lm_robust(Fertility ~ Catholic + Agriculture, data = _) |>
      tidy()
  }))

boot |>
  unnest(coefs) |>
  group_by(term) |>
  summarise(bootstrap_se = sd(estimate))

# for comparison against HC2
lm_robust(Fertility ~ Catholic + Agriculture, data = swiss) |>
  tidy() |>
  select(term, std.error)

```

The bootstrap standard errors and the HC2 standard errors are close, which is the expected result.
