---
title: "Regression tables with estimatr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression tables with estimatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE)
have <- function(...) all(vapply(c(...), requireNamespace, logical(1), quietly = TRUE))
library(estimatr)
```

An `lm_robust` fit carries its own robust standard errors, so a table package that reads the fit gets the right numbers with no second step. Two packages have methods for estimatr objects and are what this page covers: `texreg` and `modelsummary`. Both produce LaTeX, HTML and Word output from the same call.

```{r}
fit1 <- lm_robust(mpg ~ hp, data = mtcars)
fit2 <- lm_robust(mpg ~ hp + wt, data = mtcars)
fit3 <- lm_robust(mpg ~ hp + wt, data = mtcars, clusters = cyl, se_type = "stata")
```

## texreg

estimatr registers `extract` methods for `lm_robust` and `iv_robust`, so `texreg()` and its siblings work directly. Confidence intervals are shown by default, because that is what the fit reports; `include.ci = FALSE` switches to standard errors and stars.

```{r, eval = have("texreg")}
library(texreg)
texreg(list(fit1, fit2, fit3), include.ci = FALSE)
```

`htmlreg()` produces HTML and `screenreg()` prints to the console, which is the one to use while you are still deciding what goes in the table.

```{r, eval = have("texreg")}
screenreg(list(fit1, fit2, fit3), include.ci = FALSE,
          custom.model.names = c("HC2", "HC2", "CR, stata"))
```

## modelsummary

`modelsummary` reads `lm_robust` fits through `broom`-style methods, which estimatr provides.

```{r, eval = have("modelsummary")}
library(modelsummary)
modelsummary(list("HC2" = fit1, "HC2" = fit2, "CR, stata" = fit3),
             output = "markdown", gof_map = c("nobs", "r.squared"))
```

Because the standard errors already live in the fit, do not pass `vcov =` to `modelsummary` unless you actually want it to recompute them. Passing `vcov = "robust"` there overrides estimatr's HC2 with `sandwich`'s HC3 default and silently changes the table.

## What is not here

`stargazer` has no method for `lm_robust` objects. estimatr 1.x shipped `starprep()` and `commarobust()` to bridge that gap, and both are removed in 2.0; they remain as names that error and say what happened.

```{r, error = TRUE}
starprep(fit1)
```

The replacement the error names is `modelsummary`, which is the recommendation. If a paper is already formatted around stargazer, extract the pieces and hand them over instead:

```{r, eval = FALSE}
fits <- list(fit1, fit2)
stargazer::stargazer(
  lapply(fits, \(f) lm(f$call$formula, data = mtcars)),
  se = lapply(fits, \(f) f$std.error),
  p  = lapply(fits, \(f) f$p.value)
)
```

The `lm()` refit is only there to give stargazer an object it recognises; every number that appears in the table comes from the estimatr fit.
